Conceptual and methodological problems with bias detection and avoidance in natural language processing

### 2021-06-19 thesis/04-discussion.Rmd
Develop Discussion chapter
Adds more to discussion chapter

---
output:
  pdf_document: default
  html_document: default
---
# Discussion

It is worth diving deeper into some observations that arise while analyzing the results. 
<!-- Looking into specific values -->
As we pointed out, if there is no control of what cosine distance values individual word pairs have, then there is less
understanding of the exact relations between words. If the protected words that we chose do not have high similarity with harmful stereotypes, then one should consider at least two scenarios. The first one is that the choice of protected words and attributes may be corrupted. The second one is that the metric is not able to catch the hidden bias properly. In both cases it is essential to take a look at the individual values before averaging them or aggregating in other ways.

<!-- Problem with gender words --> 
One has to also remember about the specific situation present in gender dataset. Surprisingly one could observe there high cosine distances values between some female stereotypical professions and male protected words. This leads to new ideas of how to understand the bias origins and its proper detection. To start with, cosine distance seems to catch the information regarding words co-occurrence and not the semantic similarity. If a word stereotypically associated with females has high cosine distance value for male protected words, then the metric is not proper to establish the bias presence as it can be misleading. What we mean is that in same cases it can unnecessarily label words as biased while omitting other more harmful associations. What is more, this highlights again the need for control group. One could check each stereotypical attribute with each protected word but not in order to average the values. One should rather plot the distributions and investigate how the metric measures vectors geometry and hidden semantic information.

@@

<!-- Baysian method --> 
We propose the use of Bayesian method to introduce the uncertainty measure in bias detection. There are a few advantages when including this method in the bias analysis. As noted before in our example with religion dataset, one may obtain new insight into the bias issue after analyzing the mean estimates of the coeﬀicients. There seems to be a minor difference for the mean concerning the `assosiated` class. This can lead to the conclusions that we do not have strong reasons to perceive lower cosine distance as significantly related to the fact that the attribute is associated.

One may dive deeper into this idea by investigating the connection coefficients for individual words with the use of the visualization. There one observes that indeed most of the words are clustered together independently of the class that they belong to. It is of course partly due to the fact that we include uncertainty now. This example shows how understanding of the bias measurement may differ when switching to Bayesian method. One could argue that after this analysis one obtains more information on the dataset and can pursue with the further research more cautiously.

# Summary

The aim of this paper was to show the analysis of bias detection methods used in modern natural language processing projects.

In the further research it is worth to dive deeper into some issues or to examine new ideas that occurred during our work.

<!-- 
remarks:
- delete papers from github/hide them
- should we quote "debiasing" or without quoting

-->

<!-- textbf{Control group}

How to properly prepare control group in terms of quality and quantity?
<!-- + check if it may be proved in the code -->

<!--  \textbf{Verifying other similarity meaures}

Besides cosine similarity, there are other methods used to find the similarity between vectors.

a) TS-SS 
todo: check if it makes sense to calculate for all of the datasets; interpretations etc.
https://www.researchgate.net/publication/303513110_A_Hybrid_Geometric_Approach_for_Measuring_Similarity_Level_Among_Documents_and_Document_Clustering

b) WEAT test 
<!-- https://docs.responsibly.ai/word-embedding-bias.html#module-responsibly.we.weat -->

<!-- [to check https://github.com/taki0112/Vector_Similarity] -->

<!-- TODO: check in code if they make more sense -->

<!-- ## Bibliography
- https://www.amazon.com/Language-Processing-Synthesis-Lectures-Technologies/dp/1627052984/ref=as_li_ss_tl?ie=UTF8&qid=1502062931&sr=8-1&keywords=Neural+Network+Methods+in+Natural+Language+Processing&linkCode=sl1&tag=inspiredalgor-20&linkId=d63df073fea3ebe2d405820570b3ff03 ->

<!-- ## *** Additional -->

<!-- ### Man is to Computer Programmer as Woman is to Homemaker? Debiasing Word Embeddings" -->
<!-- [article https://arxiv.org/pdf/1607.06520.pdf] -->

<!-- Assumptions: -->
<!-- "We assume there is a set of gender neutral words $N\subseteq W$ such as flight attendant or shoes, which, by definition, are not specific to any gender." -->

<!-- TODO: check if it is useful for the analysis of black is .. -->

<!-- ### Identifying the bias subspace  -->

<!-- Interesting analysis about limits o subspace and research [source https://kawine.github.io/blog/nlp/2019/09/23/bias.html] -->

<!-- Resources to analyze: -->
<!-- 1. https://arxiv.org/pdf/2005.00965.pdf -->
<!-- 2. https://arxiv.org/pdf/2009.09435.pdf -->
<!-- 3. https://arxiv.org/pdf/1904.04047.pdf -->
<!-- 4. https://www.aclweb.org/anthology/P19-1166.pdf -->

<!-- ### How is cosine similarity entangled with analogy topic in NLP (is it??) -->

<!-- article:  https://arxiv.org/pdf/1905.09866.pdf -->

<!-- Critique of the usage of analogies as a proof for gender biases -->

<!-- TODO: check if there is other argument than that algorithm does not allow it -->

<!-- general thoughts 
https://www.researchgate.net/publication/349408535_Robustness_and_Reliability_of_Gender_Bias_Assessment_in_Word_Embeddings_The_Role_of_Base_Pairs

-
-
-
-->

>>>>

---
output:
  pdf_document: default
  html_document: default
---
# Discussion and summary

<!-- Bayesian method -->

We propose the use of Bayesian methods to measure uncertainty in bias detection. There are a few advantages of this method. Including uncertainty enables one to directly observe the influence of sample sizes. Analyzing individual words and connection coefficients, one may notice how `neutral` words have smaller uncertainty intervals and `different` or `associated` quite the opposite. One of the reasons for such outcome is that we used approximately 230 neutral words and only between 11-25 (the number varies from class to class) stereotypical attributes from @Manzini2019blackToCriminal article. What is more, we also pay attention to distribution and details regarding anomalous values. With the use of simple visualizations that we introduced before, we were able to indicate suspicious cosine distance values. Additionally, we compare in details how the cosine distance values and uncertainty change after the debiasing. One can verify then how the individual vectors changed and if it what was expected. Our analysis with the use of Bayesian method gave us new ideas and hypothesis concerning not only the bias detection method but also the efficiency of debiasing itself.

<!-- General insights - estimated coefficients mean -->

We created a summary table for each of the datasets: Reddit, Reddit Debiased, and Google word embeddings. Let us first analyse the general observations from estimated coefficients mean introduced in 3.6. DATASET-LEVEL COEFFICIENTS. For Google embeddings the HPDI for all classes coefficients (associated, different, human, and none) has an interval that includes zero. This can lead one to a conclusion that the impact for associated, different, human and neutral attribute is, when averaged, quite similar. This indicates how including the uncertainty may change the use and interpretation of @Manzini2019blackToCriminal MAC metric. It seems that if one focuses only on differences between means of means, it is too simplistic. In case of Reddit word embeddings the situation is similar although HPDI interval is below 0 for Gender class when looking at `associated` and `different` mean coefficient. This can suggest that there is indeed slightly stronger impact of these attributes in cosine distance being smaller. One should also notice how `associated` and `different` coefficients have quite similar HPDI interval, the highest observed absolute difference is equal to only 0.1. This suggests that again the impact of associated attributes and difference ones is not clear at first sight when looking at averaged coefficients. Finally, let's compare the HPDI intervals for Reddit and Reddit debiased datasets. For Religion and Race dataset there is a minor shift (in absolute values the highest change is equal to approximately 0.1) of the mean coefficients towards zero. However for Gender dataset there is no significant change. This is of course the general look at the data, let's now analyze individual words.

<!-- Details from the table -->

In Reddit table one can observe that for Religion class the cosine distance results for the associated attributes are for approximately 60% of the words close to neutral attributes as well. This can suggest that in some cases words concerning humans can have higher similarity with some protected words independently if they are stereotypical or neutral-human words. One should also pay attention to the fact that for all of the `associated` and `different` attributes in Religion class the uncertainty interval overlaps at some point. What is even more surprising is that for protected words, such as `torah`  `associated` attribute has the the cosine distance slightly over 1, which means no positive similarity! If the protected words that we chose do not have high similarity with harmful associated attributes, then one should consider at least three scenarios. The first one is that the choice of protected words and attributes may be corrupted. The second one is that the metric is not able to catch the hidden bias properly. The third one is that there is actually no bias between the words. Regardless of which scenario one considers, it is essential to take a look at the individual values before averaging them or aggregating in other ways. It seems that using Bayesian method can enhance the process of verifying the hypotheses concerning the choice of protected words and attributes.

Surprisingly in Gender class one can observe high cosine similarity values between some female stereotypical professions and male protected words. If a word stereotypically associated with females has low cosine distance to male protected words, then one should try to figure out the reasons for that. Cosine distance seems to capture the information on the co-occurrence of words and not on the semantic similarity strictly speaking.

In Google table ..

<!-- comments and values that refer to them: (blue and green) -->
<!-- a) google scale -->
<!-- b) associated females -->

In Reddit debiased ..

<!-- debiased: question about the change -->
<!-- - better but not for muslims -->
<!-- - corpus frequency maybe? -->
<!-- - better for male -->
<!-- - when in gender different and associated together maybe it is better - gender is very specific, words are about professions;  -->
<!-- - when in religion it is bad different and associated -->
<!-- -  in race it is not obvious - mixed sentiments  -->

Let's summarize the results of the bias detection methods analysis.

<!-- black remarks then: -->
<!-- - mess, change protected words; relative differences stay the same, by the way lipstick on a pig -->
<!-- - word lists are strange, not enough words, large HPDI -->

<!-- where to put this ? -->

<!-- summary -->

The bottom line is that if we want to take bias seriously, so should we approach the uncertainty involved in our estimations. There is no replacement for proper statistical evaluation that does not discard information about the uncertaintly involved, larger word lists are needed, and visualisation of the results for particular protected classes provides much better guidance than chasing a single metric based on a means of means.

What we mean is that in same cases it can unnecessarily label words as biased while omitting other more harmful associations. What is more, this highlights again the need for control group. One could check each stereotypical attribute with each protected word but not in order to average the values. One should rather plot the distributions and investigate how the metric measures vectors geometry and hidden semantic information.

@@

<!--

Articles:
  1. "Are We Consistently Biased? Multidimensional Analysis of Biases in Distributional Word Vectors"
  https://www.aclweb.org/anthology/S19-1010.pdf 
  2. Summary what causes bias short article
  [source https://kawine.github.io/blog/nlp/2019/09/23/bias.html]
  3. Men Also Like Shopping: Reducing Gender Bias Amplification using Corpus-level Constraint
  4. Bias in Bios: A Case Study of Semantic Representation Bias in a High-Stakes Setting
  -->

### 2021-06-19 thesis/_book/_main.tex
Adds more to discussion chapter

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}

@@

\title{Conceptual and methodological problems with bias detection and avoidance in natural language processing}
\author{Alicja Dobrzeniecka}
\date{2021-06-17}

@@

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{library}\NormalTok{(tidyverse)}
\NormalTok{muslimClass }\OtherTok{\textless{}{-}}\NormalTok{ muslim }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(protectedWord }\SpecialCharTok{==} \StringTok{"muslim"}\NormalTok{)}
\NormalTok{neutralSample }\OtherTok{\textless{}{-}} \FunctionTok{sample\_n}\NormalTok{(}\FunctionTok{filter}\NormalTok{(muslimClass,connection }\SpecialCharTok{==} \StringTok{"none"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
\NormalTok{humanSample }\OtherTok{\textless{}{-}} \FunctionTok{sample\_n}\NormalTok{(}\FunctionTok{filter}\NormalTok{(muslimClass,connection }\SpecialCharTok{==} \StringTok{"human"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
\NormalTok{muslimVis }\OtherTok{\textless{}{-}}\NormalTok{ muslimClass }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(connection }\SpecialCharTok{!=} \StringTok{"none"} \SpecialCharTok{\&}\NormalTok{ connection }\SpecialCharTok{!=}\StringTok{"human"}\NormalTok{)}
\NormalTok{muslimVis }\OtherTok{\textless{}{-}} \FunctionTok{rbind}\NormalTok{(muslimVis,neutralSample,humanSample)}
\end{Highlighting}
\end{Shaded}

@@

\hypertarget{discussion}{%
\chapter{Discussion}\label{discussion}}

>>>>

@@

\title{Conceptual and methodological problems with bias detection and avoidance in natural language processing}
\author{Alicja Dobrzeniecka}
\date{2021-06-19}

@@

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{library}\NormalTok{(tidyverse)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## -- Attaching packages --------------------------------------- tidyverse 1.3.1 --
\end{verbatim}

\begin{verbatim}
## v ggplot2 3.3.3     v purrr   0.3.4
## v tibble  3.1.2     v dplyr   1.0.6
## v tidyr   1.1.3     v stringr 1.4.0
## v readr   1.4.0     v forcats 0.5.1
\end{verbatim}

\begin{verbatim}
## -- Conflicts ------------------------------------------ tidyverse_conflicts() --
## x dplyr::filter() masks stats::filter()
## x dplyr::lag()    masks stats::lag()
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{muslimClass }\OtherTok{\textless{}{-}}\NormalTok{ muslim }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(protectedWord }\SpecialCharTok{==} \StringTok{"muslim"}\NormalTok{)}
\NormalTok{neutralSample }\OtherTok{\textless{}{-}} \FunctionTok{sample\_n}\NormalTok{(}\FunctionTok{filter}\NormalTok{(muslimClass,connection }\SpecialCharTok{==} \StringTok{"none"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
\NormalTok{humanSample }\OtherTok{\textless{}{-}} \FunctionTok{sample\_n}\NormalTok{(}\FunctionTok{filter}\NormalTok{(muslimClass,connection }\SpecialCharTok{==} \StringTok{"human"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
\NormalTok{muslimVis }\OtherTok{\textless{}{-}}\NormalTok{ muslimClass }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(connection }\SpecialCharTok{!=} \StringTok{"none"} \SpecialCharTok{\&}\NormalTok{ connection }\SpecialCharTok{!=}\StringTok{"human"}\NormalTok{)}
\NormalTok{muslimVis }\OtherTok{\textless{}{-}} \FunctionTok{rbind}\NormalTok{(muslimVis,neutralSample,humanSample)}
\end{Highlighting}
\end{Shaded}

@@

\hypertarget{discussion-and-summary}{%
\chapter{Discussion and summary}\label{discussion-and-summary}}
